Understanding models for spatial structure in small-area estimation

### 2023-05-31 src/docs_paper/paper.Rmd
Fixes to bibliography captials, some paper small improvements

```{r geometry-graph-zwe, fig.align="center", fig.cap="A map of the districts of Zimbabwe together with the corresponding adjacency graph structure $\\mathcal{G}$. \\label{fig:geometry-graph-zwe}"}
knitr::include_graphics("depends/geometry-graph-zwe.pdf")
```

Spatial structure can be encoded using a symmetric relation between areas.
Let $i \sim j$ if the areas $A_i$ and $A_j$ are adjacent or neighbouring.
Adjacency is often defined by a shared border, though other choices are also possible [@paciorek2013spatial].
The Besag model [@besag1991bayesian] is an improper conditional auto-regressive (ICAR) model where the full conditional distribution of the $i$th spatial random effect is given by
\begin{equation}
    \phi_i \, | \, \bphi_{-i} \sim \N \left(\frac{1}{n_{\delta i}} \sum_{j: j \sim i} \phi_j, \frac{1}{n_{\delta i}\tau_\phi}\right), \label{eq:besag}
\end{equation}
where $\delta i$ is the set of neighbours of $A_i$ with cardinality $n_{\delta i} = |\delta i|$ and $\bphi_{-i}$ is the vector of spatial random effects with the $i$th entry removed.
The conditional mean of the random effect $\phi_i$ is the average of its neighbours $\{\phi_j\}_{j \sim i}$ and the precision $n_{\delta i}\tau_\phi$ is proportional to the number of neighbours $n_{\delta i}$.
By Brook's lemma \cite[Chapter 2]{rue2005gaussian} the set of full conditionals of the Besag model are equivalent to the Gaussian Markov random field (GMRF) given by
\begin{equation}
    \bphi \sim \mathcal{N}(\mathbf{0}, \tau_\phi^{-1} \bR^{-}), \label{eq:gmrf}
\end{equation}
where $\bR^{-}$ is the generalised inverse of the rank-deficient structure matrix $\bR$, so-called because it defines the structure of the precision matrix, with entries
\begin{equation}
    R_{ij} =
    \begin{cases}
        n_{\delta i}, & i = j \\
        -1, & i \sim j \\
        0, & \text{otherwise.}
    \end{cases}
\end{equation}
The Markov property arises due to the conditional independence structure $p(\phi_i \, | \, \bphi_{-i}) = p(\phi_i \, | \, \bphi_{\delta i})$ whereby each area only depends on its neighbours.
This is reflected in the sparsity of $\bR$ whereby $\phi_i \perp \phi_j \, | \, \bphi_{-ij}$ if and only if $R_{ij} = 0$.
The structure matrix $\bR$ may also be expressed as the Laplacian of the adjacency graph $\mathcal{G} = (\mathcal{V}, \mathcal{E})$ with vertices $v \in \mathcal{V}$ corresponding to each area and edges $e \in \mathcal{E}$ between vertices $i$ and $j$ when $i \sim j$.
Figure \ref{fig:geometry-graph-zwe} shows the adjacency graph for the districts of Zimbabwe alongside the geometry.
In Section \ref{sec:concerns} we will discuss the appropriateness of using the graph rather than the complete geometry.

@@

The Besag model was originally proposed for use in image analysis, where areas correspond to pixels arranged in a regular lattice structure.
Since then, it has seen wider use, including in situations where the spatial structure is less regular.
We highlight a number of concerns about the model's applicability to this broader setting.
This discussion is closely linked to the modifiable areal unit problem [@openshow1979million] whereby statistical conclusions change as a result of seemingly arbitrary changes in data aggregation, and the challenge of ecological inference and the ecological fallacy [@wakefield2010aggregation], which applies particularly to inference about individuals and groups.
<!-- Missing an item about potential loss of information by using a sparse matrix, though I don't actually think this is very important. -->

@@

Summarising all spatial information with an adjacency graph represents a loss of information.
Many geometries share the same adjacency graph, and therefore the same probability model, as illustrated by Figure \ref{fig:tikz1}.
This fact is not concerning in itself, but it prompts consideration as to whether the class of geometries with the same adjacency graph is sufficiently similar to merit identical models.^[The regularity of realistic geometries may help to constrain each class to be more similar than it strictly has to be. In other words, although pathological geometries can be constructed, they are implausible in statistical practise and so not of great concern to us here.]
Intuitively the more regular the spatial structure, the less information is lost in compression to an adjacency graph.
In image analysis, very little spatial information is lost in compression a lattice structure to an adjacency graph.
On the other hand, the regions of a country, determined by political and geographic forces, tend to display greater irregularity.
As such, the appropriateness of adjacency compression varies by the type of geometry common to the application setting.

@@

In the Besag model is all adjacent areas count equally.
This assumption is unsatisfying: for most geometries, we expect some areas to be more highly correlated to some neighbours than to others.
Figure \ref{fig:tikz1} illustrates a number of heuristic features for neighbour importance, including length of shared border, and the proximity of centers of mass.
This may be remedied using more general weighted ICAR models, as we outline in Section \ref{sec:icar}.

@@

The simplest approach is to use a single point.
A natural choice is the centroid $c_i \in A_i$, given by the arithmetic mean of the latitude and longitude, which we might hope to be representative of the area.
This results in the centroid kernel
\begin{equation}
    K(A_i, A_j) = k(c_i, c_j).
\end{equation}
<!-- More broadly, given a weighting distribution $W_i$ with probability density proportional to $w_i(s)$ over $A_i$, weighted centroids could be used in the above equation. -->
The centroid kernel has been used in environmental epidemiology [@wakefield1999spatial] and to model the reproduction number of COVID-19 [@teh2021efficient].
In a model comparison study \citet[Section 3]{best2005comparison} simulated data representing heterogeneous exposure to air pollution, including elevated rates of exposure near two hypothetical point source locations, and found that the centroid kernel tended to over-smooth the high-risk areas (though it is unsuprising that a stationary covariance would struggle to recover non-stationary structure).

@@

Rather than looking to choose a single representative point, an alternative is to represent the whole area by integrating the kernel over the areas of interest.
This results in the integrated kernel
\begin{equation}
K(A_i, A_j) = \frac{1}{|A_i||A_j|} \int_{A_i} \int_{A_j} k(s, s') \text{d} s \text{d} s'. \label{eq:ik1}
\end{equation}
This covariance structure is equivalent to that obtained by aggregating a spatially continuous Gaussian process with kernel $k$ over the areal partition, studied in the machine learning literature under the name aggregated Gaussian processes [@law2018variational; @tanaka2019spatially; @yousefi2019multi; @hamelijnck2019multi].
<!-- @law2018variational develop a variational Bayesian inference method for aggregated observation models. -->
<!-- @tanaka2019spatially consider aggregation over a linear combination of latent Gaussian processes, each of which is associated to data corresponding to a different partition of the study region. -->
Unlike for the centroid kernel where $K_{ii} = 1$ for all $i$, the marginal variance of the $i$th spatial random effect $K_{ii} = K(A_i, A_i)$ varies depending on the area: becoming smaller for more compact areas and larger for areas which are of greater extent or more spread out.

@@

If available, additional information accounting for heterogeneity over $A_i$ may be incorporated into the integrated kernel.
This can be accomplished using weighting distributions $\{W_i\}$ which represent an unequal contribution of each point to the similarity measure, to give a weighted integrated kernel
\begin{equation}
K(A_i, A_j) = \frac{1}{|A_i||A_j|}\int_{A_i} \int_{A_j} w_i(s) w_j(s') k(s, s') \text{d} s \text{d} s', \label{eq:ik2}
\end{equation}
This may be useful in disease mapping, where we expect regions with populations who live close to their shared border to be more strongly correlated than regions whose populations live far apart, which could be accounted for by weighting according to a high resolution measure of population density.

@@

### Connection to log-Gaussian Cox processes

@@

### Connection to disaggregation regression

Disaggregation regression, also known as downscaling or interpolation, is another closely related approach.
Rather than focusing on the aggregate nature of areal observations as primarily a route towards better area-level estimates, disaggregation regression aims to produce high-resolution or point-level estimates from areal observations [@utazi2019spatial; @nandi2020disaggregation].
<!-- Accurate point-level estimates are close to being a necessary intermediate goal towards obtaining accurate area-level estimates. -->
<!-- Disaggregation is challenging without auxiliary information. -->
<!-- A unified view on Bayesian varying coefficient models by Franco-Villoria et al. -->

@@

The continuous ranked probability score [CRPS; @matheson1976scoring] generalises the Brier score [@brier1950verification] to distributional forecasts.
Let $\rho_i$ have posterior marginal $f(\rho_i) = p(\rho_i \, | \, y)$ and $\omega_i$ be the true value.
Writing $F$ as the cumulative distribution function corresponding to $f$ then
\begin{equation}
\text{CRPS}(f, \omega_i) = \int_{-\infty}^{\infty} (F(\rho_i) - \mathbbm{1} \{\rho_i \geq \omega_i \} )^2 \text{d}\rho_i = \int_{0}^{1} (F(\rho_i) - \mathbbm{1} \{\rho_i \geq \omega_i \} )^2 \text{d}\rho_i, \label{eq:crps}
\end{equation}
where $\mathbbm{1}$ denotes the indicator function and the second equality follows from $0 \leq \rho_i \leq 1$.
$\text{CRPS}(f, \omega_i)$ if and only if $f(\rho_i) = \delta_{\omega_i}$.
The CRPS is a strictly proper scoring rule [@gneiting2007strictly] and can be evaluated directly using samples $\{\rho_i^s\}_{s = 1}^S$ as
\begin{equation}
\text{CRPS}(f, \omega_i) \approx \frac{1}{S} \sum_{s = 1}^S | \rho_i^s - \omega_i | - \frac{1}{2S^2} \sum_{s = 1}^S \sum_{l = 1}^S | \rho_i^s - \rho_i^l |. \label{eq:crpsapprox}
\end{equation}
<!-- The logarithmic score [@good1952rational] given by $\text{LogS}(f, \omega_i) = \log f(\omega_i)$ is another popular strictly proper scoring rule, though in line with the findings of @kruger2020predictive, we found it difficult to compute satisfactorily using kernel density estimation. -->

>>>>

```{r geometry-graph-zwe, fig.align="center", fig.cap="A map of the districts of Zimbabwe together with the corresponding adjacency graph structure $\\mathcal{G}$. The Besag model operates on the adjacency graph, rather than the geometry. \\label{fig:geometry-graph-zwe}"}
knitr::include_graphics("depends/geometry-graph-zwe.pdf")
```

Spatial structure can be encoded using a symmetric relation between areas.
Let $i \sim j$ if the areas $A_i$ and $A_j$ are adjacent or neighbouring.
Adjacency is often defined by a shared border, though other choices are possible [@paciorek2013spatial].
The Besag model [@besag1991bayesian] is an improper conditional auto-regressive (ICAR) model where the full conditional distribution of the $i$th spatial random effect is given by
\begin{equation}
    \phi_i \, | \, \bphi_{-i} \sim \N \left(\frac{1}{n_{\delta i}} \sum_{j: j \sim i} \phi_j, \frac{1}{n_{\delta i}\tau_\phi}\right), \label{eq:besag}
\end{equation}
where $\delta i$ is the set of neighbours of $A_i$ with cardinality $n_{\delta i} = |\delta i|$ and $\bphi_{-i}$ is the vector of spatial random effects with the $i$th entry removed.
The conditional mean of the random effect $\phi_i$ is the average of its neighbours $\{\phi_j\}_{j \sim i}$ and the precision $n_{\delta i}\tau_\phi$ is proportional to the number of neighbours $n_{\delta i}$.
By Brook's lemma \cite[Chapter 2]{rue2005gaussian} the set of full conditionals of the Besag model are equivalent to the Gaussian Markov random field (GMRF) given by
\begin{equation}
    \bphi \sim \mathcal{N}(\mathbf{0}, \tau_\phi^{-1} \bR^{-}), \label{eq:gmrf}
\end{equation}
where $\bR^{-}$ is the generalised inverse of the rank-deficient structure matrix $\bR$, so-called because it defines the structure of the precision matrix, with entries
\begin{equation}
    R_{ij} =
    \begin{cases}
        n_{\delta i}, & i = j \\
        -1, & i \sim j \\
        0, & \text{otherwise.}
    \end{cases}
\end{equation}
The Markov property arises due to the conditional independence structure $p(\phi_i \, | \, \bphi_{-i}) = p(\phi_i \, | \, \bphi_{\delta i})$ whereby each area only depends on its neighbours.
This is reflected in the sparsity of $\bR$ such that $\phi_i \perp \phi_j \, | \, \bphi_{-ij}$ if and only if $R_{ij} = 0$.
The structure matrix $\bR$ may also be expressed as the Laplacian of the adjacency graph $\mathcal{G} = (\mathcal{V}, \mathcal{E})$ with vertices $v \in \mathcal{V}$ corresponding to each area and edges $e \in \mathcal{E}$ between vertices $i$ and $j$ when $i \sim j$.
Figure \ref{fig:geometry-graph-zwe} shows the adjacency graph for the districts of Zimbabwe alongside the geometry.
In Section \ref{sec:concerns} we will discuss the appropriateness of using the graph rather than the complete geometry.

@@

The Besag model was originally proposed for use in image analysis, where areas correspond to pixels arranged in a regular lattice structure.
Since then, it has seen wider use, including in situations where the spatial structure is less regular.
We highlight a number of concerns about the model's applicability to this broader setting.
This discussion is closely linked to the modifiable areal unit problem [@openshow1979million], whereby statistical conclusions change as a result of seemingly arbitrary changes in data aggregation, as well as the challenge of ecological inference and the ecological fallacy [@wakefield2010aggregation].
<!-- Missing an item about potential loss of information by using a sparse matrix, though I don't actually think this is very important. -->

@@

Summarising a geometry by an adjacency graph represents a loss of information.
Many geometries share the same adjacency graph, and therefore the same probability model, as illustrated by Figure \ref{fig:tikz1}.
This fact is not concerning in itself, but does prompt consideration as to whether the class of geometries with the same adjacency graph is sufficiently similar to merit identical models.^[The regularity of realistic geometries may help to constrain each class to be more similar than it strictly has to be. In other words, although pathological geometries can be constructed, they are implausible in statistical practise and so not of great concern to us here.]
Intuitively the more regular the spatial structure, the less information is lost in compression to an adjacency graph.
In image analysis, very little spatial information is lost in compression of a lattice structure to an adjacency graph.
On the other hand, the regions of a country, determined by political and geographic forces, tend to display greater irregularity.
As such, the appropriateness of adjacency compression varies by the type of geometry common to the application setting.

@@

In the Besag model is all adjacent areas count equally.
This assumption is unsatisfying: for most geometries, we expect different amounts of correlation between neighbours.
Figure \ref{fig:tikz1} illustrates a number of heuristic features for neighbour importance, including length of shared border, and the proximity of centers of mass.
More general weighted ICAR models remedy this problem, as we outline in Section \ref{sec:icar}.

@@

The simplest approach is to use a single point.
A natural choice is the centroid $c_i \in A_i$, given by the arithmetic mean of the latitude and longitude, which may be representative of the area.^[Though note that it is not guaranteed for the centroid to (even) lie within the area.]
This results in the centroid kernel
\begin{equation}
    K(A_i, A_j) = k(c_i, c_j).
\end{equation}
<!-- More broadly, given a weighting distribution $W_i$ with probability density proportional to $w_i(s)$ over $A_i$, weighted centroids could be used in the above equation. -->
The centroid kernel has been used in environmental epidemiology [@wakefield1999spatial] and to model the reproduction number of COVID-19 [@teh2021efficient].
In a model comparison study \citet[Section 3]{best2005comparison} simulated data representing heterogeneous exposure to air pollution, including elevated rates of exposure near two hypothetical point source locations, and found that the centroid kernel tended to over-smooth the high-risk areas.^[Though it is unsuprising that a stationary covariance would struggle to recover non-stationary structure.]

@@

Rather than looking to choose a single representative point, an alternative is to represent the whole area by integrating the kernel over the areas of interest.
This results in the integrated kernel
\begin{equation}
K(A_i, A_j) = \frac{1}{|A_i||A_j|} \int_{A_i} \int_{A_j} k(s, s') \text{d} s \text{d} s'. \label{eq:ik1}
\end{equation}
This covariance structure is equivalent to that obtained by aggregating a spatially continuous Gaussian process with kernel $k$ over the areal partition, and has been studied in the machine learning literature under the title aggregated Gaussian processes [@law2018variational; @tanaka2019spatially; @yousefi2019multi; @hamelijnck2019multi].
<!-- @law2018variational develop a variational Bayesian inference method for aggregated observation models. -->
<!-- @tanaka2019spatially consider aggregation over a linear combination of latent Gaussian processes, each of which is associated to data corresponding to a different partition of the study region. -->
Unlike for the centroid kernel where $K_{ii} = 1$ for all $i$, the marginal variance of the $i$th spatial random effect $K_{ii} = K(A_i, A_i)$ varies depending on the area: becoming smaller for more compact areas and larger for areas which are of greater extent or more spread out.

@@

Additional information accounting for heterogeneity over $A_i$ may be incorporated into the integrated kernel.
This can be accomplished using weighting distributions $\{W_i\}$ which represent an unequal contribution of each point to the similarity measure, to give a weighted integrated kernel
\begin{equation}
K(A_i, A_j) = \frac{1}{|A_i||A_j|}\int_{A_i} \int_{A_j} w_i(s) w_j(s') k(s, s') \text{d} s \text{d} s', \label{eq:ik2}
\end{equation}
This may be useful in disease mapping, where we expect regions with populations who live close to their shared border to be more strongly correlated than regions whose populations live far apart, which could be accounted for by weighting according to a high resolution measure of population density.

@@

### Log-Gaussian Cox processes

@@

<!-- Discuss Gary's paper here? [@konstantinoudis2020discrete] -->

### Disaggregation regression

Disaggregation regression, also known as downscaling or interpolation, is another closely related approach.
Rather than focusing on the aggregate nature of areal observations as primarily a route towards better area-level estimates, disaggregation regression aims to produce high-resolution or point-level estimates from areal observations [@utazi2019spatial; @nandi2023disaggregation].
<!-- Accurate point-level estimates are close to being a necessary intermediate goal towards obtaining accurate area-level estimates. -->
<!-- Disaggregation is challenging without auxiliary information. -->
<!-- A unified view on Bayesian varying coefficient models by Franco-Villoria et al. -->

@@

The continuous ranked probability score [CRPS; @matheson1976scoring] generalises the Brier score [@brier1950verification] to distributional forecasts.
Let $\rho_i$ have posterior marginal $f(\rho_i) = p(\rho_i \, | \, y)$ and $\omega_i$ be the true value.
Writing $F$ as the cumulative distribution function corresponding to $f$ then
\begin{equation}
\text{CRPS}(f, \omega_i) = \int_{-\infty}^{\infty} (F(\rho_i) - \mathbbm{1} \{\rho_i \geq \omega_i \} )^2 \text{d}\rho_i = \int_{0}^{1} (F(\rho_i) - \mathbbm{1} \{\rho_i \geq \omega_i \} )^2 \text{d}\rho_i, \label{eq:crps}
\end{equation}
where $\mathbbm{1}$ denotes the indicator function and the second equality follows from $0 \leq \rho_i \leq 1$.
$\text{CRPS}(f, \omega_i)$ if and only if $f(\rho_i) = \delta_{\omega_i}$.
The CRPS is a strictly proper scoring rule^[As is the logarithmic score [@good1952rational] given by $\text{LogS}(f, \omega_i) = \log f(\omega_i)$, though in line with the findings of @kruger2020predictive, we found it difficult to compute satisfactorily using kernel density estimation.] [@gneiting2007strictly] and can be evaluated directly using samples $\{\rho_i^s\}_{s = 1}^S$ as
\begin{equation}
\text{CRPS}(f, \omega_i) \approx \frac{1}{S} \sum_{s = 1}^S | \rho_i^s - \omega_i | - \frac{1}{2S^2} \sum_{s = 1}^S \sum_{l = 1}^S | \rho_i^s - \rho_i^l |. \label{eq:crpsapprox}
\end{equation}
